---
title: "eval"
author: ""
date: ""
output: html_document
---


```{r, echo=FALSE, message=FALSE}
# packages
library(ez)
library(reshape2)
library(reshape)
library(ggplot2)
library(pastecs)
library(ez)
library(data.table)
library(tidyverse)

library(dplyr)                                                  # Load dplyr package
#library(plyr)                                                   # Load plyr package
library(readr)                                                  # Load readr package
```


# Get and prepare raw data 

```{r}
library(readr)
data <- read_csv("data.csv")
```


```{r, message = FALSE, echo = FALSE}
data_excl <- subset(data,  data$desktop_conf != "1: I confirm" | data$attent_conf != "1: I confirm")
data <- subset(data, data$desktop_conf == "1: I confirm" & data$attent_conf == "1: I confirm")

```


Exclude participants whose learning accuracy (coded in "learning_yellow_accuracy" and "learning_pink_accuracy") is less than 0.875 


```{r}
data_learning_exclusion <- data %>%
  filter(learning_yellow_accuracy < 7/8 | learning_pink_accuracy < 7/8)


data <- data %>%
  filter(learning_yellow_accuracy >= 7/8 & learning_pink_accuracy >= 7/8)
```

```{r}
total_N <- nrow(data) + nrow(data_learning_exclusion)
# percentage excluded
excl_percentage <- (nrow(data_learning_exclusion) / total_N) * 100
excl_percentage
```

```{r}
table(data$device_mechanism, data$contingency)
table(data$condition)
```
# Demographics 

```{r}
# mean age of participants 
mean(data$age, na.rm = TRUE)
# age range of participants 
range(data$age, na.rm = TRUE)
```
the one participant who indicated an age of 3 years wrote in the "tech_issues" field that they are actually 34 years old. So we will recode this participant's age to 34.

```{r}
data$age <- ifelse(data$age == 3, 34, data$age)
```


run the demographics code again

```{r}
# mean age of participants 
mean(data$age, na.rm = TRUE)
# age range of participants 
range(data$age, na.rm = TRUE)
# number of genders reported 
table(data$gender)
```


# Data preparation 

Make factors

```{r}
data$contingency <- factor(data$contingency, 
                              levels = c("1_1", "05_05", "0_0"), 
                              labels = c("P(e+|c+) 4/4 \nP(e+|c-) 4/4", "P(e+|c+) 2/4 \nP(e+|c-) 2/4", 
                                         "P(e+|c+) 0/4 \nP(e+|c-) 0/4"))


data$device_mechanism <- factor(data$device_mechanism, 
                              levels = c("opaque", "transp"), 
                              labels = c("opaque", "transparent"))
```



# Analysis


## Plots 

Make a data frame in long format for plotting prior and posterior structure ratings together

```{r}
# make a separate data frame in long format for plotting, with the DVs being "structure_rating_prior" and "strucutre_rating_posterior"
data_long <- data %>%
  pivot_longer(cols = c(structure_rating_prior, structure_rating_posterior),
               names_to = "causal_rating_type",
               values_to = "rating_value")

# make rating_type a factor with levels in the desired order
data_long$causal_rating_type <- factor(data_long$causal_rating_type, levels = c("structure_rating_prior", "structure_rating_posterior"),
                                labels = c("Prior", "Posterior"))
```





```{r}
library(dplyr)
library(ggplot2)
library(Hmisc)  # for mean_cl_boot



device_mechanism_colors <- c(
  # Dark2 Colors (Colors 1-8)
  "opaque"      = "#1b9e77",  # Dark2 Green
  "transparent"   = "#7570b3"  # Dark2 Purple
)



plotDV <- function(df, dv_col, x_jitter = 0.10, y_jitter = 5) {
  
  
  ggplot(df, aes(x = causal_rating_type, y = .data[[dv_col]], fill = device_mechanism)) +
    facet_grid(device_mechanism ~ contingency) +
    scale_y_continuous(limits = c(-10, 110), breaks = seq(0, 100, by = 10)) +
    
    # Reference line at y = 50
    geom_hline(yintercept = 50, color = "grey20", linewidth = 0.4, linetype = "dashed") +
    
    # Violin background (lighter, no outline)
    geom_violin(
      trim = TRUE, alpha = 0.25,
      color = NA
    ) +
    
    # Subject lines (jittered both horizontally & vertically)
    geom_line(
      aes(group = subj_code),
      alpha = 0.1,
      color = "grey50",
      position = position_jitter(width = x_jitter, height = y_jitter)
    ) +
    
    # Jittered points
    geom_point(
      alpha = 0.15,
      size = 1.2,
      color = "black",
      position = position_jitter(width = x_jitter, height = y_jitter)
    ) +
    
    # Bootstrapped mean ±95% CI ribbons (higher alpha)
    stat_summary(
      aes(group = device_mechanism, fill = device_mechanism),
      fun.data = mean_cl_boot,
      geom = "ribbon",
      alpha = 0.55,
      color = NA
    ) +
    
    # Mean line (per device_mechanism)
    stat_summary(
      aes(color = device_mechanism, group = device_mechanism),
      fun = mean,
      geom = "line",
      size = 1
    ) +
    
    # Mean point with thin black border
    stat_summary(
      aes(fill = device_mechanism),
      fun.data = mean_cl_boot,
      geom = "pointrange",
      shape = 21,
      color = "black",
      size = 0.9,
      stroke = 0.6
    ) +
    
    # Median point (white triangle)
    stat_summary(
      fun = median,
      geom = "point",
      shape = 24,
      size = 2.5,
      fill = "white",
      color = "black"
    ) +
    
    # Color scales
    scale_fill_manual(values = device_mechanism_colors) +
    scale_color_manual(values = device_mechanism_colors) +
    
    labs(
      x = "Structure Rating",
      y = "Rating Value"
    ) +
    
    # APA-friendly theme (no gridlines)
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 25, hjust = 1),
      axis.text.y = element_text(size = 10),
      panel.spacing = unit(3, "lines"),
      panel.grid = element_blank(),
      strip.background = element_rect(fill = "white", color = "grey70", linewidth = 0.4),
      strip.text = element_text(face = "bold", size = 12),
      axis.title = element_text(face = "bold")
    )
}

# ---- Run and Save ----
p1 <- plotDV(data_long, "rating_value")
p1

#ggsave("structure_rating_by_device_mechanism_APA_colored.svg", plot = p1, width = 8, height = 6)
#ggsave("structure_rating_by_device_mechanism_APA_colored.pdf", plot = p1, width = 8, height = 6)


```

```{r}
# get the median ratings in each condition 
data_long %>%
  group_by(device_mechanism, contingency, causal_rating_type) %>%
  summarise(median_rating = median(rating_value)) -> medians

medians
```



## Statistical tests

Test the difference between prior and posterior structure ratings depending on device_mechanism and contingency condition using a mixed ANOVA with contingency and device_mechanism as between-subject factors and rating_type (prior vs. posterior) as within-subject factor


```{r}
library(afex)
library(emmeans)

anova_data <- data_long

a1 <- aov_car(rating_value ~ device_mechanism*causal_rating_type*contingency  + Error(subj_code/causal_rating_type), anova_data, anova_table = list(es = "pes")) # the anova model
a1 
```

Repeat analysis as mixed effects model (same results in different format)

```{r}
library(lme4)
library(lmerTest)  # gives p-values

# drop the non-realized contingency levels 
anova_data <- droplevels(anova_data)
with(anova_data, table(device_mechanism, contingency))

model <- lmer(
  rating_value ~ device_mechanism * causal_rating_type * contingency +
    (1 | subj_code),
  data = anova_data
)

anova(model, type = 3)
```

List of the group means 

```{r}
# means
group_means <- emmeans(a1, c("causal_rating_type", "contingency", "device_mechanism")) 
group_means
```

compute a contrast with emmeans to test if the prior-posterior change is larger in the opaque device_mechanism condition compared to the transparent device_mechanism condition, across all contingency conditions

```{r}
# main effect of device_mechanism
pairs(emmeans(a1, ~ device_mechanism), adjust = "none")

# main effect of rating type
pairs(emmeans(a1, ~ causal_rating_type), adjust = "none")


#
pairs(emmeans(a1, ~ device_mechanism*causal_rating_type), adjust = "none")


# rating differences across device_mechanism conditions
pairs(emmeans(a1, ~ causal_rating_type|device_mechanism), adjust = "none")

# now test the difference in these deltas between the two device_mechanism conditions
pairs(pairs(emmeans(a1, ~ causal_rating_type*device_mechanism), adjust = "none"), adjust = "none")

```

```{r}
pairs(emmeans(a1, ~ causal_rating_type|contingency*device_mechanism), adjust = "none") -> deltas

deltas
```



## Rating change categories

Compute the different between prior and posterior for each participant and append it to data frame 

```{r}
data <- data %>%
  mutate(rating_diff = structure_rating_posterior - structure_rating_prior)
```

Code for each participate what kind of rating difference occurred: no difference, increase, decrease. For the "no difference" type, allow for some noise of plus minus 5 points 

```{r}
data <- data %>%
  mutate(rating_diff_type = case_when(
    rating_diff > 0 ~ "increase",
    rating_diff < 0 ~ "decrease",
    TRUE ~ "same"
  ))
```


Now make a plot that shows the proportions of these rating_diff_type for each device_mechanism and contingency condition; as error bars, show 95% Proportion CIs (using binom.confint from the binom package)

```{r}
library(binom)
data_diff_type <- data %>%
  group_by(device_mechanism, contingency, rating_diff_type) %>%
  summarise(
    count = n()
  ) %>%
  group_by(device_mechanism, contingency) %>%
  mutate(
    total = sum(count),
    proportion = count / total
  ) %>%
  ungroup()
# compute 95% CIs for proportions
data_diff_type <- data_diff_type %>%
  rowwise() %>%
  mutate(
    ci = list(binom.confint(count, total, method = "wilson"))
  ) %>%
  unnest_wider(ci) %>%
  select(-c(method, mean, x))


ggplot(data_diff_type, aes(x = contingency, y = proportion, fill = rating_diff_type)) +
  facet_wrap(~ device_mechanism) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.7) +
  geom_errorbar(aes(ymin = lower, ymax = upper), 
                position = position_dodge(width = 0.9), width = 0.2) +
  scale_fill_manual(values = c("increase" = "#1b9e77", "same" = "#d95f02", "decrease" = "#7570b3")) +
  labs(
    title = "Proportion of Rating Change Types",
    x = "Contingency",
    y = "Proportion", 
    fill = "change type"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 25, hjust = 1)
  )+ # define y axis limits so that they go from 0 to 1 in steps of 0.25
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25))-> p3
p3
# save plot 
#ggsave("subgroups.svg", plot = p3, width = 14, height = 6)
#ggsave("subgroups.pdf", plot = p3, width = 14, height = 6)
```

Determine the proportion of participants in the "transparent" condition who fell into the "same" cluster and a "rating_diff" of 0
```{r}
data %>%
  filter(device_mechanism == "transparent" & rating_diff_type == "same") %>%
  summarise(
    count = n()
  ) -> transparent_same_count

data %>%
  filter(device_mechanism == "transparent" & rating_diff_type == "same" & structure_rating_prior == 0) %>%
  summarise(
    count = n()
  ) -> transparent_same_count_zero_prior


transparent_same_count_zero_prior/transparent_same_count
```

In the transparent conditions, of the participants who gave identical prior and posterior ratings, 82% had a prior rating of 0.



# Generate Bayesian model predictions 

```{r}
# number of simulations 

m = 100000
```


## SI model 

```{r}
# --- Load function file ---
source("SI_model_fun.R")

# --- Set up contingency tables ---
data <- matrix(data = 0, nrow = 5, ncol = 4)
#             D  C  B  A              
data[1,] <- c(0, 4, 0, 4) # 1.0 1.0
data[2,] <- c(1, 3, 1, 3) # 0.75 0.75
data[3,] <- c(2, 2, 2, 2) # 0.5 0.5
data[4,] <- c(3, 1, 3, 1) # 0.25 0.25
data[5,] <- c(4, 0, 4, 0) # 0.0 0.0

# --- Sweep over priors ---
prior_grid <- seq(0, 1, by = 0.1)

# Store all results in a list
all_preds <- list()

for (p0 in prior_grid) {
  p1 <- 1 - p0
  message("Running model with prior_S0 = ", p0, ", prior_S1 = ", p1)
  
  preds <- generate_SI_preds(
    data = data,
    m = m,
    prior_S0 = p0,
    prior_S1 = p1
  )
  
  # Add prior info as columns for later identification
  preds$prior_S0 <- p0
  preds$prior_S1 <- p1
  
  # Store results
  all_preds[[as.character(p0)]] <- preds
}

# Combine all into one big data frame
library(dplyr)
pred_SI_all <- bind_rows(all_preds, .id = "prior_S0_label")

# make a subset containing columns 2:5, S0_pp, S1_pp, prior_S0, prior_S1
pred_SI_all <- pred_SI_all %>%
  select(N_00, N_01, N_10, N_11, S0_pp, S1_pp, prior_S0, prior_S1)


pred_SI_all$contingency <- pred_SI_all$N_11



pred_SI_all$contingency <- factor(pred_SI_all$contingency, 
                              levels = c("4", "3", "2", "1", "0"), 
                              labels = c("P(e+|c+) 4/4 \nP(e+|c-) 4/4", "P(e+|c+) 3/4 \nP(e+|c-) 3/4", "P(e+|c+) 2/4 \nP(e+|c-) 2/4", 
                                         "P(e+|c+) 1/4 \nP(e+|c-) 1/4", "P(e+|c+) 0/4 \nP(e+|c-) 0/4"))

```

```{r}
library(ggplot2)

ggplot(pred_SI_all, aes(x = prior_S1, y = S1_pp, color = contingency, group = contingency)) +
  # 45° reference line
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40", linewidth = 1) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_x_continuous(
    name = expression("Prior: P(S"[1]*")"),
    breaks = seq(0, 1, 0.1),
    limits = c(0, 1)
  ) +
  scale_y_continuous(
    name = expression("Posterior: P(S"[1]*" | data)"),
    breaks = seq(0, 1, 0.1),
    limits = c(0, 1)
  ) +
  scale_color_viridis_d(option = "C", end = 0.9) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    legend.title = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  ) +
  ggtitle(expression("Posterior belief in S1 as a function of prior belief in S1")) +
  coord_equal()+
  theme(
    plot.margin = margin(t = 0, r = 0, b = 0, l = 0)
  )

# save plot as svg 

#ggsave("SI_model_prior_vs_posterior_S1_N8.svg", width = 7, height = 5)
#ggsave("SI_model_prior_vs_posterior_S1_N8.pdf", width = 7, height = 5)
```

```{r}

library(ggnewscale)

library(dplyr)
library(tidyr)
library(ggplot2)
library(grid)

# --- Prepare data ---
seg <- pred_SI_all %>%
  select(contingency, prior_S1, S1_pp) %>%
  transmute(
    contingency,
    id   = prior_S1,
    x0   = 0,   # Prior of S1
    x1   = 1,   # Posterior of S1
    y0   = prior_S1,
    y1   = S1_pp
  )

# Highlighted priors
seg_hi <- seg %>%
  filter(near(id, 0) | near(id, 0.5)) %>%
  mutate(highlight_label = ifelse(near(id, 0), "Prior = 0", "Prior = 0.5"))

# Endpoints for points + labels
endpts_hi <- seg_hi %>%
  tidyr::pivot_longer(c(x0, x1, y0, y1),
                      names_to = c(".value","which"),
                      names_pattern = "([xy])(\\d)") %>%
  mutate(
    which = ifelse(which == "0", "Prior of S1", "Posterior of S1"),
    x = ifelse(which == "Prior of S1", 0, 1),
    y = y
  )

# --- Plot ---
ggplot() +
  # Reference line at y = 50
    geom_hline(yintercept = 0.50, color = "grey20", linewidth = 0.4, linetype = "dashed") +
  # faint background arrows
  geom_segment(
    data = seg,
    aes(x = x0, xend = x1, y = y0, yend = y1),
    color = "black",
    linewidth = 0.8, alpha = 0.2,
    arrow = arrow(length = unit(0.15, "cm"), type = "closed")
  ) +

  # highlighted arrows (two distinct colors)
  geom_segment(
    data = seg_hi,
    aes(x = x0, xend = x1, y = y0, yend = y1, color = highlight_label),
    linewidth = 2.5, alpha = 0.85,
    arrow = arrow(length = unit(0.25, "cm"), type = "closed")
  ) +
  #geom_point(data = endpts_hi,aes(x = x, y = y, color = highlight_label),size = 3, alpha = 0.9) +
  geom_text(data = endpts_hi,
            aes(x = x, y = y, label = sprintf("%.2f", y)),
            vjust = -1.1, size = 3, color = "black") +

  facet_wrap(~contingency, nrow = 1) +
  scale_color_manual(
    name = "Highlighted priors",
    values = c(
      "Prior = 0"   = "#7570b3",  # red-orange #1b9e77", "transparent" = "#7570b3
      "Prior = 0.5" = "#1b9e77"   # teal
    ),
    guide = guide_legend(
      title.position = "top",
      override.aes = list(linewidth = 3, alpha = 1)
    )
  ) +
  scale_x_continuous(
    breaks = c(0, 1),
    labels = c(expression("Prior of S"[1]),
               expression("Post. of S"[1])),
    limits = c(-0.1, 1.1),
    name = NULL
  ) +
  scale_y_continuous(
    name = "Probability",
    limits = c(0, 1),
    breaks = seq(0, 1, 0.1)
  ) +
  theme_minimal(base_size = 14) +
  theme(
  # …your other theme settings…
  panel.grid.major.x = element_blank(),
  panel.grid.minor.x = element_blank(),
  panel.grid.minor.y = element_blank(),
  panel.grid.major.y = element_blank(),
  panel.spacing.x = unit(1.5, "lines"),   # <-- NEW LINE: adds space between facets
  legend.position = "right",
  legend.title = element_text(size = 12, face = "bold"),
  legend.text = element_text(size = 10),
  strip.text = element_text(face = "bold", size = 12),
  plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
  axis.text.x = element_text(size = 11,angle = 25, hjust = 1)
)+
  labs(
    title = expression("Prior–Posterior Updates for S"[1]*" across Contingency Conditions")
  )-> g_update_highlight

g_update_highlight

# Save plot
#ggsave("SI_model_prior_posterior_updates_highlight.svg", plot = g_update_highlight, width = 14, height = 5)

```


```{r}
library(ggnewscale)
library(dplyr)
library(tidyr)
library(ggplot2)
library(grid)

# --- Subset contingencies of interest (8/8, 4/4, 0/0) ---
subset_data <- pred_SI_all %>%
  filter(N_11 %in% c(4, 2, 0))

# --- Prepare data ---
seg <- subset_data %>%
  select(contingency, prior_S1, S1_pp) %>%
  transmute(
    contingency,
    id   = prior_S1,
    x0   = 0,   # Prior of S1
    x1   = 1,   # Posterior of S1
    y0   = prior_S1,
    y1   = S1_pp
  )

# Highlighted priors
seg_hi <- seg %>%
  filter(near(id, 0) | near(id, 0.5)) %>%
  mutate(highlight_label = ifelse(near(id, 0), "Prior = 0", "Prior = 0.5"))

# Endpoints for points + labels
endpts_hi <- seg_hi %>%
  tidyr::pivot_longer(c(x0, x1, y0, y1),
                      names_to = c(".value","which"),
                      names_pattern = "([xy])(\\d)") %>%
  mutate(
    which = ifelse(which == "0", "Prior of S1", "Posterior of S1"),
    x = ifelse(which == "Prior of S1", 0, 1),
    y = y
  )

# --- Plot ---
ggplot() +
  geom_hline(yintercept = 0.50, color = "grey20", linewidth = 0.4, linetype = "dashed") +
  geom_segment(
    data = seg,
    aes(x = x0, xend = x1, y = y0, yend = y1),
    color = "black",
    linewidth = 0.8, alpha = 0.2,
    arrow = arrow(length = unit(0.15, "cm"), type = "closed")
  ) +
  geom_segment(
    data = seg_hi,
    aes(x = x0, xend = x1, y = y0, yend = y1, color = highlight_label),
    linewidth = 2.5, alpha = 0.85,
    arrow = arrow(length = unit(0.25, "cm"), type = "closed")
  ) +
  geom_text(data = endpts_hi,
            aes(x = x, y = y, label = sprintf("%.2f", y)),
            vjust = -1.1, size = 3, color = "black") +
  facet_wrap(~contingency, nrow = 1) +
  scale_color_manual(
    name = "Highlighted priors",
    values = c(
      "Prior = 0"   = "#7570b3",
      "Prior = 0.5" = "#1b9e77"
    ),
    guide = guide_legend(
      title.position = "top",
      override.aes = list(linewidth = 3, alpha = 1)
    )
  ) +
  scale_x_continuous(
    breaks = c(0, 1),
    labels = c(expression("Prior of S"[1]),
               expression("Post. of S"[1])),
    limits = c(-0.1, 1.1),
    name = NULL
  ) +
  scale_y_continuous(
    name = "Probability",
    limits = c(0, 1),
    breaks = seq(0, 1, 0.1)
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.spacing.x = unit(2, "lines"),
    legend.position = "right",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    strip.text = element_text(face = "bold", size = 12),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text.x = element_text(size = 11, angle = 0, hjust = 0.5)
  ) +
  labs(
    title = expression("Prior–Posterior Updates for S"[1]*"")
  ) -> g_update_highlight_subset

g_update_highlight_subset

# Save subset plot
#ggsave("SI_model_prior_posterior_updates_highlight_subset_N8.svg",
#       plot = g_update_highlight_subset, width = 10, height = 4)

#ggsave("SI_model_prior_posterior_updates_highlight_subset_N8.pdf",
#       plot = g_update_highlight_subset, width = 10, height = 4)


```

