---
title: "Analysis"
format: html
---



```{r, echo=FALSE, message=FALSE}
# packages
library(ez)
library(reshape2)
library(reshape)
library(ggplot2)
library(pastecs)
library(ez)
library(data.table)
library(tidyverse)

library(dplyr)                                                  # Load dplyr package
#library(plyr)                                                   # Load plyr package
library(readr)                                                  # Load readr package
```


# Get and prepare raw data 

```{r}
library(readr)
data <- read_csv("data.csv")
```


```{r, message = FALSE, echo = FALSE}
data_excl <- subset(data,  data$desktop_conf != "1: I confirm" | data$attent_conf != "1: I confirm")
data <- subset(data, data$desktop_conf == "1: I confirm" & data$attent_conf == "1: I confirm")

```


Exclude participants whose learning accuracy (coded in "learning_yellow_accuracy" and "learning_pink_accuracy") is less than 0.875 


```{r}
data_learning_exclusion <- data %>%
  filter(learning_yellow_accuracy < 6/8 | learning_pink_accuracy < 6/8)


data <- data %>%
  filter(learning_yellow_accuracy >= 6/8 & learning_pink_accuracy >= 6/8)
```

```{r}
total_N <- nrow(data) + nrow(data_learning_exclusion)
# percentage excluded
excl_percentage <- (nrow(data_learning_exclusion) / total_N) * 100
excl_percentage
```

```{r}
table(data$prior_level, data$contingency, data$prior_type)
table(data$condition)
```

# Demographics 

```{r}
# make column "age" numeric 
data$age <- as.numeric(data$age)
# mean age of participants 
mean(data$age, na.rm = TRUE)
# age range of participants 
range(data$age, na.rm = TRUE)

table(data$gender)
```

# Data preparation 

Make factors

```{r}
data$contingency <- factor(data$contingency, 
                              levels = c("1_1", "05_05"), 
                              labels = c("P(e+|c+) = 4/4 \nP(e+|c-) = 4/4",
                                         "P(e+|c+) = 2/4 \nP(e+|c-) = 2/4"))



data$prior_level <- factor(data$prior_level, 
                              levels = c("informative_S0", "informative_S1"), 
                              labels = c("suggests S0", "suggests S1"))


data$prior_type <- factor(data$prior_type, 
                              levels = c("mechanism", "statistical"), 
                              labels = c("mechanistic", "statistical"))
```



# Analysis


## Plots 

Make a data frame in long format for plotting prior and posterior structure ratings together

```{r}
# make a separate data frame in long format for plotting, with the DVs being "structure_rating_prior" and "strucutre_rating_posterior"
data_long <- data %>%
  pivot_longer(cols = c(rating_prior, rating_posterior),
               names_to = "causal_rating_type",
               values_to = "rating_value")

# make rating_type a factor with levels in the desired order
data_long$causal_rating_type <- factor(data_long$causal_rating_type, levels = c("rating_prior", "rating_posterior"),
                                labels = c("Prior", "Posterior"))
```



```{r}
library(dplyr)
library(ggplot2)
library(Hmisc)  # for mean_cl_boot

# ---- Setup Colors ----
prior_level_colors <- c(
  "suggests S0" = "#7570b3", 
  "suggests S1" = "#d95f02"
)

plotDV_consistent <- function(df, dv_col, x_jitter = 0.12, y_jitter = 4) {
  
  # --- METHOD 2: PRE-CALCULATE MATCHED JITTER COORDINATES ---
  df_jittered <- df %>%
    mutate(x_numeric = as.numeric(as.factor(causal_rating_type))) %>%
    # Group by subject and all grouping variables to lock the line trajectory shifts on the X-axis
    group_by(subj_code, prior_level, contingency, prior_type) %>%
    mutate(
      x_shift = runif(1, -x_jitter, x_jitter)
    ) %>%
    ungroup() %>%
    # Group by row to assign an individual Y-shift that lines and points will share perfectly
    rowwise() %>%
    mutate(
      y_shift = runif(1, -y_jitter, y_jitter),
      x_jittered = x_numeric + x_shift,
      y_jittered = .data[[dv_col]] + y_shift
    ) %>%
    ungroup()

  # Extract original phase labels for axis formatting
  x_labels <- levels(as.factor(df$causal_rating_type))
  
  # --- BUILD PLOT WITH ALIGNED JITTER ---
  ggplot(df_jittered, aes(x = x_jittered, y = y_jittered, 
                          fill = prior_level, color = prior_level)) +
    
    # Setting as.table = FALSE swaps the row order (plotting from bottom-up)
    facet_grid(prior_level ~ contingency + prior_type, as.table = FALSE) + 
    
    # Map text labels back onto the continuous baseline positions
    scale_x_continuous(breaks = seq_along(x_labels), labels = x_labels) +
    scale_y_continuous(limits = c(-5, 105), breaks = seq(0, 100, by = 25)) +
    scale_fill_manual(values = prior_level_colors) +
    scale_color_manual(values = prior_level_colors) +
    
    geom_hline(yintercept = 50, color = "grey30", linewidth = 0.5, linetype = "dotted") +
    
    # Background Distributions (Using true un-jittered variables)
    geom_violin(aes(x = as.numeric(as.factor(causal_rating_type)), y = .data[[dv_col]], group = causal_rating_type), 
                trim = TRUE, alpha = 0.45, color = NA) +
    
    # Individual Subject Trajectories
    geom_line(
      aes(group = subj_code),
      alpha = 0.15,
      color = "grey50"
    ) +
    
    # Raw Data Points
    geom_point(
      alpha = 0.12,
      size = 1,
      color = "black"
    ) +
    
    # Medians (Using true un-jittered variables to keep them centered and accurate)
    stat_summary(
      aes(x = as.numeric(as.factor(causal_rating_type)), y = .data[[dv_col]]),
      fun = median,              
      geom = "point",          
      shape = 21,              
      color = "black",         
      size = 2.5,              
      stroke = 1.0, 
      fill = "white"
    ) +
    
    labs(
      x = "Rating Phase",
      y = "Structure Rating (0–100)", 
      fill = "Prior Information", 
      color = "Prior Information"
    ) +
    
    theme_classic(base_size = 12) +
    theme(
      legend.title = element_text(face = "bold"),
      legend.position = "bottom", 
      axis.text.x = element_text(size = 10),
      axis.line = element_line(linewidth = 0.5),
      panel.spacing = unit(1.2, "lines"),
      strip.background = element_blank(),
      strip.text.x = element_text(face = "bold", size = 12),
      strip.text.y = element_blank(),                       
      plot.margin = margin(10, 15, 10, 10)
    )
}

# ---- Run and Save ----
p_participants <- plotDV_consistent(data_long, "rating_value")
p_participants

if(!dir.exists("Figures")) dir.create("Figures")
ggsave("Figures/Participants.svg", plot = p_participants, width = 8, height = 6, device = "svg")
```



```{r}
library(dplyr)
library(ggplot2)
library(Hmisc)  # for mean_cl_boot


# ---- Setup Colors ----
prior_level_colors <- c(
  "suggests S0" = "#7570b3", 
  "suggests S1" = "#d95f02"
)

# ---- Plotting Function: Publication Consistent (Prior vs. Posterior) ----
plotDV_consistent <- function(df, dv_col, x_jitter = 0.12, y_jitter = 4) {
  
  ggplot(df, aes(x = causal_rating_type, y = .data[[dv_col]], 
                 fill = prior_level, color = prior_level)) +
    
    # Facet grid with explicit labels for the Mechanism rows
    facet_grid(
       ~ contingency+prior_type, 
    ) + 
    
    # Scale and Aesthetics
    scale_y_continuous(limits = c(-5, 105), breaks = seq(0, 100, by = 25)) +
    scale_fill_manual(values = prior_level_colors) +
    scale_color_manual(values = prior_level_colors) +
    
    # Reference line at chance/neutrality
    geom_hline(yintercept = 50, color = "grey30", linewidth = 0.5, linetype = "dotted") +
    
    # 4. Bootstrapped 95% CI Ribbons
    stat_summary(
      aes(group = prior_level),
      fun.data = mean_cl_boot,
      geom = "ribbon",
      alpha = 0.4,
      color = NA
    ) +
    
    # 5. Mean Trajectory Line
    stat_summary(
      aes(group = prior_level),
      fun = mean,
      geom = "line",
      linewidth = 1 # Slightly thicker for emphasis
    ) +
    
# 6. Mean Point (Simple point instead of pointrange)
    stat_summary(
      fun = mean,              # Just calculate the mean
      geom = "point",          # Use a simple point geom
      shape = 21,              # Fillable circle
      color = "black",         # Outline color
      size = 2.5,              # Size of the point
      stroke = 1.0             # Thickness of the outline
    ) +
    
    # Labels
    labs(
      x = "Rating Phase",
      y = "Structure Rating (0–100)", 
      fill = "Prior Information", 
      color = "Prior Information"
    ) +
    
    # Consistent Publication Theme
    theme_classic(base_size = 12) +
    theme(
      legend.title = element_text(face = "bold"),
      legend.position = "bottom", 
      axis.text.x = element_text(size = 10),
      axis.line = element_line(linewidth = 0.5),
      panel.spacing = unit(1.2, "lines"),
      strip.background = element_blank(),
      # --- KEY CHANGES HERE ---
      strip.text.x = element_text(face = "bold", size = 12), # Keep contingency labels
      strip.text.y = element_blank(),                       # Hide mechanism labels
      # -------------------------
      plot.margin = margin(10, 15, 10, 10)
    )
}

# ---- Run and Save ----
p_means <- plotDV_consistent(data_long, "rating_value")
p_means
# Save as high-res PDF and PNG
if(!dir.exists("Figures")) dir.create("Figures")

# also as svg
ggsave("Figures/Exp4_Means.svg", plot = p_means, width = 7, height = 4, device = "svg")
# save also as pdf 
ggsave("Figures/Exp4_Means.pdf", plot = p_means, width = 7, height = 4, device = "pdf")
```

```{r}
# Install patchwork if you haven't yet
# install.packages("patchwork")
library(patchwork)

# Combine the plots
# / signifies putting p_means below p_participants
# | signifies putting them side-by-side
combined_plot <- p_participants | p_means

# Add labels and fine-tune
final_plot <- combined_plot + 
  plot_annotation(tag_levels = 'A') & 
  theme(plot.tag = element_text(face = "bold", size = 16))

# Display the plot
final_plot

# Save the combined figure
ggsave("Figures/Exp4_Combined_Figure.svg", plot = final_plot, width = 12.5, height = 5, device = "svg")
# save also as pdf 
ggsave("Figures/Exp4_Combined_Figure.pdf", plot = final_plot, width = 12.5, height = 5, device = "pdf")
```



```{r}
# get the median ratings in each condition 
data_long %>%
  group_by(prior_level, prior_type, contingency, causal_rating_type) %>%
  summarise(median_rating = median(rating_value)) -> medians

medians
```



## Statistical tests

Test the difference between prior and posterior structure ratings depending on device_mechanism and contingency condition using a mixed ANOVA with contingency and device_mechanism as between-subject factors and rating_type (prior vs. posterior) as within-subject factor


```{r}
library(afex)
library(emmeans)

anova_data <- data_long

a1 <- aov_car(rating_value ~ prior_level*prior_type*causal_rating_type*contingency  + Error(subj_code/causal_rating_type), anova_data, anova_table = list(es = "pes")) # the anova model
a1 
```

Repeat analysis as mixed effects model (same results in different format)

```{r}
library(lme4)
library(lmerTest)  # gives p-values

# drop the non-realized contingency levels 
#anova_data <- droplevels(anova_data)
#with(anova_data, table(device_mechanism, contingency))

model <- lmer(
  rating_value ~ prior_level * prior_type * causal_rating_type * contingency +
    (1 | subj_code),
  data = anova_data
)

anova(model, type = 3)
```

List of the group means 

```{r}
# means
group_means <- emmeans(a1, c("causal_rating_type", "prior_level", "prior_type", "contingency")) 
group_means
```

compute a contrast with emmeans to test if the prior-posterior change is larger in the opaque device_mechanism condition compared to the transparent device_mechanism condition, across all contingency conditions

```{r}
# main effect of device_mechanism
pairs(emmeans(a1, ~ prior_level), adjust = "none")
```


```{r}
# main effect of rating type
pairs(emmeans(a1, ~ causal_rating_type), adjust = "none")
```

```{r}
# rating differences across device_mechanism conditions
pairs(emmeans(a1, ~ causal_rating_type|prior_level), adjust = "none")
```



```{r}
pairs(emmeans(a1, ~ causal_rating_type|contingency*prior_level), adjust = "none") -> deltas

deltas
```




# Subset 

Now we make subsets: in the conditions suggesting S1, we keep participants with values around of at least 70 and in the conditions suggesting S0, we keep participants with values around not higher than 30. 

```{r}
library(dplyr)

data_sub <- data %>%
  filter(
    (prior_level == "suggests S0" & rating_prior <= 30) |
    (prior_level == "suggests S1" & rating_prior >= 70)
  )
```




## Plots 

Make a data frame in long format for plotting prior and posterior structure ratings together

```{r}
# make a separate data frame in long format for plotting, with the DVs being "structure_rating_prior" and "strucutre_rating_posterior"
data_long <- data_sub %>%
  pivot_longer(cols = c(rating_prior, rating_posterior),
               names_to = "causal_rating_type",
               values_to = "rating_value")

# make rating_type a factor with levels in the desired order
data_long$causal_rating_type <- factor(data_long$causal_rating_type, levels = c("rating_prior", "rating_posterior"),
                                labels = c("Prior", "Posterior"))
```



```{r}
library(dplyr)
library(ggplot2)
library(Hmisc)  # for mean_cl_boot

# ---- Setup Colors ----
prior_level_colors <- c(
  "suggests S0" = "#7570b3", 
  "suggests S1" = "#d95f02"
)

plotDV_consistent <- function(df, dv_col, x_jitter = 0.12, y_jitter = 4) {
  
  # --- METHOD 2: PRE-CALCULATE MATCHED JITTER COORDINATES ---
  df_jittered <- df %>%
    mutate(x_numeric = as.numeric(as.factor(causal_rating_type))) %>%
    # Group by subject and all active conditions to lock line shifts together on the X-axis
    group_by(subj_code, prior_level, contingency, prior_type) %>%
    mutate(
      x_shift = runif(1, -x_jitter, x_jitter)
    ) %>%
    ungroup() %>%
    # Group by individual rows to create matching Y-shifts for both lines and points
    rowwise() %>%
    mutate(
      y_shift = runif(1, -y_jitter, y_jitter),
      x_jittered = x_numeric + x_shift,
      y_jittered = .data[[dv_col]] + y_shift
    ) %>%
    ungroup()

  # Extract original factor levels for labeling the X-axis
  x_labels <- levels(as.factor(df$causal_rating_type))
  
  # --- BUILD PLOT WITH ALIGNED JITTER ---
  ggplot(df_jittered, aes(x = x_jittered, y = y_jittered, 
                          fill = prior_level, color = prior_level)) +
    
    # Setting as.table = FALSE swaps the row order (plotting from bottom-up)
    facet_grid(prior_level ~ contingency + prior_type, as.table = FALSE) + 
    
    # Map text labels back onto the continuous baseline positions
    scale_x_continuous(breaks = seq_along(x_labels), labels = x_labels) +
    scale_y_continuous(limits = c(-5, 105), breaks = seq(0, 100, by = 25)) +
    scale_fill_manual(values = prior_level_colors) +
    scale_color_manual(values = prior_level_colors) +
    
    geom_hline(yintercept = 50, color = "grey30", linewidth = 0.5, linetype = "dotted") +
    
    # Background Distributions (Using true un-jittered variables)
    geom_violin(aes(x = as.numeric(as.factor(causal_rating_type)), y = .data[[dv_col]], group = causal_rating_type), 
                trim = TRUE, alpha = 0.45, color = NA) +
    
    # Individual Subject Trajectories
    geom_line(
      aes(group = subj_code),
      alpha = 0.15,
      color = "grey50"
    ) +
    
    # Raw Data Points
    geom_point(
      alpha = 0.12,
      size = 1,
      color = "black"
    ) +
    
    # Medians (Using true un-jittered variables to keep them centered and accurate)
    stat_summary(
      aes(x = as.numeric(as.factor(causal_rating_type)), y = .data[[dv_col]]),
      fun = median,              
      geom = "point",          
      shape = 21,              
      color = "black",         
      size = 2.5,              
      stroke = 1.0, 
      fill = "white"
    ) +
    
    labs(
      x = "Rating Phase",
      y = "Structure Rating (0–100)", 
      fill = "Prior Information", 
      color = "Prior Information"
    ) +
    
    theme_classic(base_size = 12) +
    theme(
      legend.title = element_text(face = "bold"),
      legend.position = "bottom", 
      axis.text.x = element_text(size = 10),
      axis.line = element_line(linewidth = 0.5),
      panel.spacing = unit(1.2, "lines"),
      strip.background = element_blank(),
      strip.text.x = element_text(face = "bold", size = 12),
      strip.text.y = element_blank(),                       
      plot.margin = margin(10, 15, 10, 10)
    )
}

# ---- Run and Save ----
p_participants <- plotDV_consistent(data_long, "rating_value")
p_participants

if(!dir.exists("Figures")) dir.create("Figures")
ggsave("Figures/Participants_Sub.svg", plot = p_participants, width = 8, height = 6, device = "svg")
```



```{r}
library(dplyr)
library(ggplot2)
library(Hmisc)  # for mean_cl_boot


# ---- Setup Colors ----
prior_level_colors <- c(
  "suggests S0" = "#7570b3", 
  "suggests S1" = "#d95f02"
)

# ---- Plotting Function: Publication Consistent (Prior vs. Posterior) ----
plotDV_consistent <- function(df, dv_col, x_jitter = 0.12, y_jitter = 4) {
  
  ggplot(df, aes(x = causal_rating_type, y = .data[[dv_col]], 
                 fill = prior_level, color = prior_level)) +
    
    # Facet grid with explicit labels for the Mechanism rows
    facet_grid(
       ~ contingency+prior_type, 
    ) + 
    
    # Scale and Aesthetics
    scale_y_continuous(limits = c(-5, 105), breaks = seq(0, 100, by = 25)) +
    scale_fill_manual(values = prior_level_colors) +
    scale_color_manual(values = prior_level_colors) +
    
    # Reference line at chance/neutrality
    geom_hline(yintercept = 50, color = "grey30", linewidth = 0.5, linetype = "dotted") +
    
    # 4. Bootstrapped 95% CI Ribbons
    stat_summary(
      aes(group = prior_level),
      fun.data = mean_cl_boot,
      geom = "ribbon",
      alpha = 0.4,
      color = NA
    ) +
    
    # 5. Mean Trajectory Line
    stat_summary(
      aes(group = prior_level),
      fun = mean,
      geom = "line",
      linewidth = 1 # Slightly thicker for emphasis
    ) +
    
# 6. Mean Point (Simple point instead of pointrange)
    stat_summary(
      fun = mean,              # Just calculate the mean
      geom = "point",          # Use a simple point geom
      shape = 21,              # Fillable circle
      color = "black",         # Outline color
      size = 2.5,              # Size of the point
      stroke = 1.0             # Thickness of the outline
    ) +
    
    # Labels
    labs(
      x = "Rating Phase",
      y = "Structure Rating (0–100)", 
      fill = "Prior Information", 
      color = "Prior Information"
    ) +
    
    # Consistent Publication Theme
    theme_classic(base_size = 12) +
    theme(
      legend.title = element_text(face = "bold"),
      legend.position = "bottom", 
      axis.text.x = element_text(size = 10),
      axis.line = element_line(linewidth = 0.5),
      panel.spacing = unit(1.2, "lines"),
      strip.background = element_blank(),
      # --- KEY CHANGES HERE ---
      strip.text.x = element_text(face = "bold", size = 12), # Keep contingency labels
      strip.text.y = element_blank(),                       # Hide mechanism labels
      # -------------------------
      plot.margin = margin(10, 15, 10, 10)
    )
}

# ---- Run and Save ----
p_means <- plotDV_consistent(data_long, "rating_value")
p_means
# Save as high-res PDF and PNG
if(!dir.exists("Figures")) dir.create("Figures")

# also as svg
ggsave("Figures/Exp4_Means_Sub.svg", plot = p_means, width = 7, height = 4, device = "svg")
# save also as pdf 
ggsave("Figures/Exp4_Means_Sub.pdf", plot = p_means, width = 7, height = 4, device = "pdf")
```

```{r}
# Install patchwork if you haven't yet
# install.packages("patchwork")
library(patchwork)

# Combine the plots
# / signifies putting p_means below p_participants
# | signifies putting them side-by-side
combined_plot <- p_participants | p_means

# Add labels and fine-tune
final_plot <- combined_plot + 
  plot_annotation(tag_levels = 'A') & 
  theme(plot.tag = element_text(face = "bold", size = 16))

# Display the plot
final_plot

# Save the combined figure
ggsave("Figures/Exp4_Combined_Figure_Sub.svg", plot = final_plot, width = 12.5, height = 5, device = "svg")
# save also as pdf 
ggsave("Figures/Exp4_Combined_Figure_Sub.pdf", plot = final_plot, width = 12.5, height = 5, device = "pdf")
```


# Repeat the statistical analysis with the subset of participants

```{r}
library(afex)
library(emmeans)

anova_data <- data_long

a1 <- aov_car(rating_value ~ prior_level*prior_type*causal_rating_type*contingency  + Error(subj_code/causal_rating_type), anova_data, anova_table = list(es = "pes")) # the anova model
a1 
```

Repeat analysis as mixed effects model (same results in different format)

```{r}
library(lme4)
library(lmerTest)  # gives p-values

# drop the non-realized contingency levels 
#anova_data <- droplevels(anova_data)
#with(anova_data, table(device_mechanism, contingency))

model <- lmer(
  rating_value ~ prior_level * prior_type * causal_rating_type * contingency +
    (1 | subj_code),
  data = anova_data
)

anova(model, type = 3)
```


