---
title: "Analysis"
format: html
---

```{r, echo=FALSE, message=FALSE}
# Standard Tidyverse Stack
library(tidyverse)   # Loads dplyr, ggplot2, readr, tidyr, etc. cleanly
library(data.table)

# Modeling & Statistics
library(afex)
library(emmeans)
library(lme4)
library(lmerTest)

# Plotting Utilities & Specialized Stats
library(patchwork)
library(Hmisc)       # Required for mean_cl_boot
library(binom)       # For proportion CIs

# Legacy/Reshaping & Descriptive Utilities 
# Note: Loaded in specific order to avoid package masking conflicts
library(reshape)
library(reshape2)
library(pastecs)
library(ez)          

# -------------------------------------------------------------------------
# SET GLOBAL SEED FOR REPRODUCIBILITY
# -------------------------------------------------------------------------
# This locks down both the simulation maths and the ggplot jitter coordinates
set.seed(42)
```

# Get and prepare raw data

```{r}
# Read in raw participant data (Assumes a local "data.csv")
data <- read_csv("data.csv")
```

```{r, message = FALSE, echo = FALSE}
# Exclusion based on confirmation prompts
data_excl <- subset(data, data$desktop_conf != "1: I confirm" | data$attent_conf != "1: I confirm")
data <- subset(data, data$desktop_conf == "1: I confirm" & data$attent_conf == "1: I confirm")

# Exclude participants whose learning accuracy is less than 0.875
data_learning_exclusion <- data %>%
  filter(learning_yellow_accuracy < 7/8 | learning_pink_accuracy < 7/8)

data <- data %>%
  filter(learning_yellow_accuracy >= 7/8 & learning_pink_accuracy >= 7/8)

# Calculate exclusion percentages
total_N <- nrow(data) + nrow(data_learning_exclusion)
excl_percentage <- (nrow(data_learning_exclusion) / total_N) * 100
message("Excluding ", round(excl_percentage, 2), "% of participants based on learning accuracy.")
```

```{r}
table(data$device_mechanism, data$contingency)
table(data$condition)
```

# Demographics

```{r}
data$age <- as.numeric(data$age)
mean(data$age, na.rm = TRUE)
range(data$age, na.rm = TRUE)
```

# Data preparation

Make factors

```{r}
data$contingency <- factor(data$contingency, 
                           levels = c("1_1", "05_05", "0_0"), 
                           labels = c("P(e+|c+) = 4/4 \nP(e+|c-) = 4/4",
                                      "P(e+|c+) = 2/4 \nP(e+|c-) = 2/4", 
                                      "P(e+|c+) = 0/4 \nP(e+|c-) = 0/4"))

data$device_mechanism <- factor(data$device_mechanism, 
                                 levels = c("opaque", "transp"), 
                                 labels = c("uninformative", "suggests S0"))
```

# Analysis

## Plots

Make a data frame in long format for plotting prior and posterior structure ratings together

```{r}
# Create long-format dataset for plotting prior and posterior structure ratings together
data_long <- data %>%
  pivot_longer(cols = c(structure_rating_prior, structure_rating_posterior),
               names_to = "causal_rating_type",
               values_to = "rating_value")

data_long$causal_rating_type <- factor(data_long$causal_rating_type, 
                                       levels = c("structure_rating_prior", "structure_rating_posterior"),
                                       labels = c("Prior", "Posterior"))

data_exp <- data
```


```{r}
set.seed(42)
# ---- Setup Colors ----
device_mechanism_colors <- c(
  "suggests S0"   = "#7570b3",  
  "suggests S1"   = "#d95f02",
  "uninformative" = "#1b9e77"   
)

plotDV_consistent_m2 <- function(df, dv_col, x_jitter = 0.12, y_jitter = 4) {
  df_jittered <- df %>%
    mutate(x_numeric = as.numeric(as.factor(causal_rating_type))) %>%
    group_by(subj_code, device_mechanism, contingency) %>%
    mutate(x_shift = runif(1, -x_jitter, x_jitter)) %>%
    ungroup() %>%
    rowwise() %>%
    mutate(
      y_shift = runif(1, -y_jitter, y_jitter),
      x_jittered = x_numeric + x_shift,
      y_jittered = .data[[dv_col]] + y_shift
    ) %>%
    ungroup()

  x_labels <- levels(as.factor(df$causal_rating_type))

  ggplot(df_jittered, aes(x = x_jittered, y = y_jittered, 
                          fill = device_mechanism, color = device_mechanism)) +
    facet_grid(device_mechanism ~ contingency) + 
    scale_x_continuous(breaks = seq_along(x_labels), labels = x_labels) +
    scale_y_continuous(limits = c(-5, 105), breaks = seq(0, 100, by = 25)) +
    scale_fill_manual(values = device_mechanism_colors) +
    scale_color_manual(values = device_mechanism_colors) +
    geom_hline(yintercept = 50, color = "grey30", linewidth = 0.5, linetype = "dotted") +
    geom_violin(aes(x = as.numeric(as.factor(causal_rating_type)), y = .data[[dv_col]], group = causal_rating_type), 
                trim = TRUE, alpha = 0.45, color = NA) +
    geom_line(aes(group = subj_code), alpha = 0.15, color = "grey50") +
    geom_point(alpha = 0.12, size = 1, color = "black") +
    stat_summary(
      aes(x = as.numeric(as.factor(causal_rating_type)), y = .data[[dv_col]]),
      fun = median,              
      geom = "point",          
      shape = 21,              
      color = "black",         
      size = 2.5,              
      stroke = 1.0, 
      fill = "white"
    ) +
    labs(
      x = "Rating Phase",
      y = "Structure Rating (0–100)", 
      fill = "Device Mechanism", 
      color = "Device Mechanism"
    ) +
    theme_classic(base_size = 12) +
    theme(
      legend.title = element_text(face = "bold"),
      legend.position = "bottom", 
      axis.text.x = element_text(size = 10),
      axis.line = element_line(linewidth = 0.5),
      panel.spacing = unit(1.2, "lines"),
      strip.background = element_blank(),
      strip.text.x = element_text(face = "bold", size = 12), 
      strip.text.y = element_blank(),                       
      plot.margin = margin(10, 15, 10, 10)
    )
}

p_participants_m2 <- plotDV_consistent_m2(data_long, "rating_value", x_jitter = 0.12, y_jitter = 4)
p_participants_m2
ggsave("Figures/Participants_m2.svg", plot = p_participants_m2, width = 6, height = 6, device = "svg")
```



```{r}
set.seed(42)
plotDV_means <- function(df, dv_col) {
  ggplot(df, aes(x = causal_rating_type, y = .data[[dv_col]], 
                 fill = device_mechanism, color = device_mechanism)) +
    facet_grid(~ contingency) + 
    scale_y_continuous(limits = c(-5, 105), breaks = seq(0, 100, by = 25)) +
    scale_fill_manual(values = device_mechanism_colors) +
    scale_color_manual(values = device_mechanism_colors) +
    geom_hline(yintercept = 50, color = "grey30", linewidth = 0.5, linetype = "dotted") +
    stat_summary(
      aes(group = device_mechanism),
      fun.data = mean_cl_boot,
      geom = "ribbon",
      alpha = 0.4,
      color = NA
    ) +
    stat_summary(
      aes(group = device_mechanism),
      fun = mean,
      geom = "line",
      linewidth = 1 
    ) +
    stat_summary(
      fun = mean,              
      geom = "point",          
      shape = 21,              
      color = "black",         
      size = 2.5,              
      stroke = 1.0 
    ) +
    labs(
      x = "Rating Phase",
      y = "Structure Rating (0–100)", 
      fill = "Device Mechanism", 
      color = "Device Mechanism"
    ) +
    theme_classic(base_size = 12) +
    theme(
      legend.title = element_text(face = "bold"),
      legend.position = "bottom", 
      axis.text.x = element_text(size = 10),
      axis.line = element_line(linewidth = 0.5),
      panel.spacing = unit(1.2, "lines"),
      strip.background = element_blank(),
      strip.text.x = element_text(face = "bold", size = 12), 
      strip.text.y = element_blank(),                       
      plot.margin = margin(10, 15, 10, 10)
    )
}

p_means <- plotDV_means(data_long, "rating_value")
ggsave("Figures/Means.svg", plot = p_means, width = 5.5, height = 4, device = "svg")
```

```{r}
combined_plot <- p_participants_m2 | p_means
final_plot <- combined_plot + 
  plot_annotation(tag_levels = 'A') & 
  theme(plot.tag = element_text(face = "bold", size = 16))

combined_plot

ggsave("Figures/Exp1_Combined_Figure.svg", plot = final_plot, width = 12, height = 5, device = "svg")
ggsave("Figures/Exp1_Combined_Figure.pdf", plot = final_plot, width = 12, height = 5, device = "pdf")
```

```{r}
baseSize = 16
clean_theme <- theme_classic(base_size = baseSize) +  
  theme(
    panel.background = element_blank(),
    plot.background = element_blank(),
    strip.background = element_blank(),
    axis.line = element_line(color = "black")
  )

plot_priors <- function(bc, wc, wa) {
  df <- data.frame(x = seq(0, 1, length.out = 300))
  
  p1 <- ggplot(df, aes(x = x)) +
    geom_area(stat = "function", fun = dbeta, args = list(shape1 = bc[1], shape2 = bc[2]), fill = "forestgreen", alpha = 0.2) +
    stat_function(fun = dbeta, args = list(shape1 = bc[1], shape2 = bc[2]), color = "forestgreen", size = 1) +
    labs(title = bquote("Prior for b"["c"]*": beta(" * .(bc[1]) * "," * .(bc[2]) * ")"), x = "Value", y = "Density") + clean_theme

  p2 <- ggplot(df, aes(x = x)) +
    geom_area(stat = "function", fun = dbeta, args = list(shape1 = wc[1], shape2 = wc[2]), fill = "orange", alpha = 0.2) +
    stat_function(fun = dbeta, args = list(shape1 = wc[1], shape2 = wc[2]), color = "orange", size = 1) +
    labs(title = bquote("Prior for w"["c"]*": beta(" * .(wc[1]) * "," * .(wc[2]) * ")"), x = "Value", y = "Density") + clean_theme

  p3 <- ggplot(df, aes(x = x)) +
    geom_area(stat = "function", fun = dbeta, args = list(shape1 = wa[1], shape2 = wa[2]), fill = "firebrick", alpha = 0.2) +
    stat_function(fun = dbeta, args = list(shape1 = wa[1], shape2 = wa[2]), color = "firebrick", size = 1) +
    labs(title = bquote("Prior for w"["a"]*": beta(" * .(wa[1]) * "," * .(wa[2]) * ")"), x = "Value", y = "Density") + clean_theme

  wrap_elements(full = (p1 | p2 | p3))
}
```

```{r}
# Master wrapper function to avoid copy-pasting the heavy loop block 4 times
source("SI_model_fun.R")
set.seed(42)

run_si_simulation <- function(bc_prior, wc_prior, wa_prior, label_text) {
  m = 500000 
  sim_data <- matrix(data = c(2,2,2,2, 4,0,4,0, 0,4,0,4), nrow = 3, ncol = 4, byrow = TRUE)
  prior_grid <- seq(0, 1, by = 0.05)
  all_preds <- list()

  for (p0 in prior_grid) {
    p1 <- 1 - p0
    preds <- generate_SI_preds(data = sim_data, m = m, prior_S0 = p0, prior_S1 = p1,
                               bc_prior = bc_prior, wc_prior = wc_prior, wa_prior = wa_prior)
    preds$prior_S0 <- p0
    preds$prior_S1 <- p1
    all_preds[[as.character(p0)]] <- preds
  }

  res <- bind_rows(all_preds, .id = "prior_S0_label") %>%
    select(N_00, N_01, N_10, N_11, S0_pp, S1_pp, prior_S0, prior_S1) %>%
    mutate(contingency = case_when(
      N_00 == 2 & N_01 == 2 & N_10 == 2 & N_11 == 2 ~ "P(e+|c+) = 2/4 \nP(e+|c-) = 2/4",
      N_00 == 4 & N_01 == 0 & N_10 == 4 & N_11 == 0 ~ "P(e+|c+) = 0/4 \nP(e+|c-) = 0/4",
      N_00 == 0 & N_01 == 4 & N_10 == 0 & N_11 == 4 ~ "P(e+|c+) = 4/4 \nP(e+|c-) = 4/4",
      TRUE ~ paste(N_00, N_01, N_10, N_11, sep = "-")
    )) %>%
    select(S1_pp, prior_S1, contingency) %>%
    mutate(contingency = factor(contingency, levels = c("P(e+|c+) = 4/4 \nP(e+|c-) = 4/4",
                                                        "P(e+|c+) = 2/4 \nP(e+|c-) = 2/4",
                                                        "P(e+|c+) = 0/4 \nP(e+|c-) = 0/4"))) %>%
    filter(round(prior_S1, 2) %in% c(0.5, 0.10))
  
  res$parameter_priors <- label_text
  return(res)
}
```

```{r}
set.seed(42)
# 1. Flat Model
distr_flat <- plot_priors(c(1,1), c(1,1), c(1,1))
pred_SI_subset_flat <- run_si_simulation(c(1,1), c(1,1), c(1,1), "flat")

# 2. Mech informed 050
distr_inf_050 <- plot_priors(c(80,80), c(1,10), c(80,80))
pred_SI_subset_inf_050 <- run_si_simulation(c(80,80), c(1,10), c(80,80), "mech. informed") %>%
  filter(contingency == "P(e+|c+) = 2/4 \nP(e+|c-) = 2/4")

# 3. Mech informed 10
distr_inf_10 <- plot_priors(c(80,80), c(1,10), c(40,5))
pred_SI_subset_inf_10 <- run_si_simulation(c(80,80), c(1,10), c(4,5), "mech. informed") %>%
  filter(contingency == "P(e+|c+) = 4/4 \nP(e+|c-) = 4/4")

# 4. Mech informed 0
distr_inf_0 <- plot_priors(c(80,80), c(1,10), c(1,10))
pred_SI_subset_inf_0 <- run_si_simulation(c(80,80), c(1,10), c(1,10), "mech. informed") %>%
  filter(contingency == "P(e+|c+) = 0/4 \nP(e+|c-) = 0/4")

# 5. Strong wc & weak wa
distr_inf_ss <- plot_priors(c(80,80), c(40,5), c(1,10))
pred_SI_subset_inf_ss <- run_si_simulation(c(80,80), c(40,5), c(1,10), "strong wc \n& weak wa")

# Bind it all up
pred_SI <- rbind(pred_SI_subset_inf_050, pred_SI_subset_inf_10, pred_SI_subset_inf_0, pred_SI_subset_inf_ss, pred_SI_subset_flat)

pred_SI <- pred_SI %>%
  mutate(device_mechanism = case_when(
    abs(prior_S1 - 0.5) < 1e-5 ~ "uninformative",
    abs(prior_S1 - 0.1) < 1e-5 ~ "suggests S0",
    TRUE                       ~ NA_character_
  ))

pred_SI_long <- pred_SI %>%
  pivot_longer(cols = c(prior_S1, S1_pp), names_to = "causal_rating_type", values_to = "prediction") %>%
  mutate(causal_rating_type = case_match(causal_rating_type, "prior_S1" ~ "Prior", "S1_pp" ~ "Posterior", .default = causal_rating_type))

pred_SI_long$causal_rating_type <- factor(pred_SI_long$causal_rating_type, levels = c("Prior", "Posterior"))
pred_SI_long$device_mechanism <- factor(pred_SI_long$device_mechanism, levels = c("uninformative", "suggests S0"))



# create plot
plot_model_predictions <- function(df) {
  ggplot(df, aes(x = causal_rating_type, y = prediction, 
                 group = interaction(device_mechanism, parameter_priors), 
                 color = device_mechanism, fill = device_mechanism)) +
    facet_grid(~ contingency) + 
    scale_y_continuous(limits = c(-0.05, 1.05), breaks = seq(0, 1, by = 0.25)) +
    scale_fill_manual(values = device_mechanism_colors) +
    scale_color_manual(values = device_mechanism_colors) +
    scale_linetype_manual(values = c("mech. informed" = "dashed", 
                                     "strong wc \n& weak wa" = "dotdash", 
                                     "flat" = "dotted")) +
    geom_hline(yintercept = 0.5, color = "grey30", linewidth = 0.5, linetype = "dotted") +
    geom_line(aes(linetype = parameter_priors), linewidth = 1) +
    geom_point(shape = 21, color = "black", size = 2.5, stroke = 1.0) +
    labs(
      x = "Rating Phase",
      y = expression("Model Prediction: P(S"[1]*" | D)"),
      fill = "Device Mechanism", 
      color = "Device Mechanism",
      linetype = "Parameter Priors"
    ) +
    theme_classic(base_size = 12) +
    theme(
      legend.title = element_text(face = "bold"),
      legend.position = "right", 
      legend.box = "vertical", 
      axis.text.x = element_text(size = 10),
      axis.line = element_line(linewidth = 0.5),
      panel.spacing = unit(1.2, "lines"),
      strip.background = element_blank(),
      strip.text.x = element_text(face = "bold", size = 12), 
      plot.margin = margin(10, 15, 10, 10)
    )
}

p_predictions <- plot_model_predictions(pred_SI_long) +
  guides(
    linetype = guide_legend(override.aes = list(linewidth = 0.8), keywidth = unit(2.5, "lines")),
    color = guide_legend(keywidth = unit(1.5, "lines")),
    fill = guide_legend(override.aes = list(shape = 21, size = 3))
  )


# put the graphs together
library(patchwork)

# Combine the plots: Row 1 has two plots, Row 2 has the model predictions and an empty space
combined_plot <- (p_participants_m2 | p_means) / (p_predictions | plot_spacer())

# Add layout annotations and styles
final_plot <- combined_plot + 
  plot_annotation(tag_levels = 'A') & 
  theme(plot.tag = element_text(face = "bold", size = 20))

# Display the plot
final_plot

# Save the combined figure
#ggsave("Figures/Exp1_Combined_Figure_with_Pred.svg", plot = final_plot, width = 12, height = 10, device = "svg")
# save also as pdf 
#ggsave("Figures/Exp1_Combined_Figure_with_Pred.pdf", plot = final_plot, width = 12, height = 10, device = "pdf")
```

```{r}
library(dplyr)

data_means <- data_long %>%
  group_by(device_mechanism, contingency, causal_rating_type) %>%
  dplyr::summarize(
    mean_rating = mean(rating_value, na.rm = TRUE),
    .groups = "drop"
  )
```

Model fits:

```{r}
data_means <- data_long %>%
  group_by(device_mechanism, contingency, causal_rating_type) %>%
  dplyr::summarize(mean_rating = mean(rating_value, na.rm = TRUE), .groups = "drop")

fit_df <- pred_SI_long %>%
  inner_join(data_means, by = c("contingency", "device_mechanism", "causal_rating_type")) %>%
  mutate(prediction_scaled = prediction * 100)

fit_metrics_by_mechanism <- fit_df %>%
  group_by(device_mechanism, parameter_priors) %>%
  dplyr::summarize(
    RMSE = sqrt(mean((mean_rating - prediction_scaled)^2, na.rm = TRUE)),
    Correlation = cor(mean_rating, prediction_scaled, use = "complete.obs"),
    R_Squared = Correlation^2,
    .groups = "drop"
  ) %>%
  arrange(device_mechanism, RMSE)

print(fit_metrics_by_mechanism)
```


## Statistical tests

Test the difference between prior and posterior structure ratings depending on device_mechanism and contingency condition using a mixed ANOVA with contingency and device_mechanism as between-subject factors and rating_type (prior vs. posterior) as within-subject factor

```{r}
anova_data <- data_long
a1 <- aov_car(rating_value ~ device_mechanism * causal_rating_type * contingency  + Error(subj_code/causal_rating_type), anova_data, anova_table = list(es = "pes"))
print(a1)
```

Repeat analysis as mixed effects model (same results in different format)

```{r}
model <- lmer(rating_value ~ device_mechanism * causal_rating_type * contingency + (1 | subj_code), data = anova_data)
anova(model, type = 3)
```

List of the group means

```{r}
# means
group_means <- emmeans(a1, c("causal_rating_type", "device_mechanism", "contingency")) 
group_means
```

compute a planned contrasts with emmeans

```{r}
group_means <- emmeans(a1, c("causal_rating_type", "device_mechanism", "contingency")) 
print(group_means)

pairs(emmeans(a1, ~ device_mechanism), adjust = "none") # main effect 
pairs(emmeans(a1, ~ causal_rating_type), adjust = "none") # main effect 
pairs(emmeans(a1, ~ causal_rating_type|device_mechanism), adjust = "none")

deltas <- pairs(emmeans(a1, ~ causal_rating_type|contingency*device_mechanism), adjust = "none")
print(deltas)
```



## Rating change categories

Compute the different between prior and posterior for each participant and append it to data frame

```{r}
data <- data_exp %>%
  mutate(
    rating_diff = structure_rating_posterior - structure_rating_prior,
    rating_diff_type = case_when(
      rating_diff > 0 ~ "P(S1) < P(S1|D)",
      rating_diff < 0 ~ "P(S1) > P(S1|D)",
      TRUE ~ "P(S1) = P(S1|D)"
    )
  )
```

Now make a plot that shows the proportions of these rating_diff_type for each device_mechanism and contingency condition; as error bars, show 95% Proportion CIs (using binom.confint from the binom package)

```{r}
data_diff_type <- data %>%
  group_by(device_mechanism, contingency, rating_diff_type) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(device_mechanism, contingency) %>%
  mutate(total = sum(count), proportion = count / total) %>%
  ungroup() %>%
  rowwise() %>%
  mutate(ci = list(binom.confint(count, total, method = "wilson"))) %>%
  unnest_wider(ci) %>%
  select(-c(method, mean, x))

p3 <- ggplot(data_diff_type, aes(x = contingency, y = proportion, fill = rating_diff_type)) +
  facet_wrap( ~ device_mechanism) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.7) +
  geom_errorbar(aes(ymin = lower, ymax = upper), position = position_dodge(width = 0.9), width = 0.2) +
  scale_fill_manual(values = c("P(S1) < P(S1|D)" = "#1b9e77", "P(S1) = P(S1|D)" = "#d95f02", "P(S1) > P(S1|D)" = "#7570b3")) +
  labs(x = "Contingency", y = "Proportion", fill = "Change Type") +
  theme_minimal(base_size = 20) +
  theme(legend.position = "top", axis.text.x = element_text(angle = 25, hjust = 1)) + 
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25))

p3

ggsave("Figures/Exp1_subgroups.svg", plot = p3, width = 14, height = 6)
ggsave("Figures/Exp1_subgroups.pdf", plot = p3, width = 14, height = 6)
```

