---
title: "eval"
author: ""
date: ""
output: html_document
---


```{r, echo=FALSE, message=FALSE}
# packages
library(ez)
library(reshape2)
library(reshape)
library(ggplot2)
library(pastecs)
library(ez)
library(data.table)
library(tidyverse)

library(dplyr)                                                  # Load dplyr package
#library(plyr)                                                   # Load plyr package
library(readr)                                                  # Load readr package
```



# Get and prepare raw data 



```{r, message = FALSE, echo = FALSE}
library(readr)
data <- read_csv("data.csv")

data_excl <- subset(data,  data$desktop_conf != "1: I confirm" | data$attent_conf != "1: I confirm")
data <- subset(data, data$desktop_conf == "1: I confirm" & data$attent_conf == "1: I confirm")

```



```{r}
table(data$machine, data$contingency)
```


Exclude participants whose learning accuracy (coded in "learning_yellow_accuracy" and "learning_pink_accuracy") is less than 0.875 


```{r}
data_learning_exclusion <- data %>%
  filter(learning_yellow_accuracy < 7/8 | learning_pink_accuracy < 7/8)


data <- data %>%
  filter(learning_yellow_accuracy >= 7/8 & learning_pink_accuracy >= 7/8)
```

```{r}
total_N <- nrow(data) + nrow(data_learning_exclusion)
# percentage excluded
excl_percentage <- (nrow(data_learning_exclusion) / total_N) * 100
excl_percentage
```

```{r}
table(data$machine, data$contingency)
table(data$condition)
```


```{r}
# exclude last participants in the 0_0 conditions (because we aimed for N = 50 in each condition)
library(dplyr)

data <- data %>%
  arrange(machine, contingency, recorded_at) %>%   # earliest -> latest within condition
  group_by(machine, contingency) %>%
  slice_head(n = 50) %>%                           # keep first 50 per condition
  ungroup()


table(data$machine, data$contingency)
table(data$condition)
```



# Analysis

```{r}



# relable factor "str_query_type" so that "prob" becomes "agree" 

data$contingency <- factor(data$contingency, 
                              levels = c("1_1", "075_075", "05_05", "025_025", "0_0"), 
                              labels = c("P(e+|c+) 4/4 \nP(e+|c-) 4/4", "P(e+|c+) 3/4 \nP(e+|c-) 3/4", "P(e+|c+) 2/4 \nP(e+|c-) 2/4", 
                                         "P(e+|c+) 1/4 \nP(e+|c-) 1/4", "P(e+|c+) 0/4 \nP(e+|c-) 0/4"))


data$machine <- factor(data$machine, 
                              levels = c("opaque", "transp"), 
                              labels = c("opaque", "transparent"))



table(data$machine, data$contingency)



```


```{r}
descriptives_prior <- data %>%
  group_by(machine, contingency) %>%
  summarise(
    mean_confidence = mean(structure_rating_prior, na.rm = TRUE),
    sd_confidence   = sd(structure_rating_prior, na.rm = TRUE)
  )

descriptives_prior

```

```{r}
descriptives_post <- data %>%
  group_by(machine, contingency) %>%
  summarise(
    mean_confidence = mean(structure_rating_posterior, na.rm = TRUE),
    sd_confidence   = sd(structure_rating_posterior, na.rm = TRUE)
  )

descriptives_post
```

# Plots 

first a plot that shows only the posterior structure ratings for each machine and contingency condition, with individual data points, mean and 95% CI

```{r}
machine_colors <- c(
  # Dark2 Colors (Colors 1-8)
  "opaque"      = "#1b9e77",  # Dark2 Green
  "1.0 \n0.5"   = "#d95f02",  # Dark2 Orange
  "transparent"   = "#7570b3",  # Dark2 Purple
  "0.25 \n0.25" = "#e7298a",  # Dark2 Pink/Magenta
  "0.75 \n0.75" = "#66a61e",  # Dark2 Light Green
  "0.625 \n0.5" = "#e6ab02",  # Dark2 Yellow/Brown
  "0.875 \n0.5" = "#a6761d"   # Dark2 Brown (The added color)
  
  # Note: The 8th Dark2 color is #666666 (Dark Gray), which you could use if you had an 8th condition.
)
```





Now make a data frame in long format for plotting prior and posterior structure ratings together

```{r}
# make a separate data frame in long format for plotting, with the DVs being "structure_rating_prior" and "strucutre_rating_posterior"
data_long <- data %>%
  pivot_longer(cols = c(structure_rating_prior, structure_rating_posterior),
               names_to = "rating_type",
               values_to = "rating_value")

# make rating_type a factor with levels in the desired order
data_long$rating_type <- factor(data_long$rating_type, levels = c("structure_rating_prior", "structure_rating_posterior"),
                                labels = c("Prior", "Posterior"))
```





```{r}
library(dplyr)
library(ggplot2)
library(Hmisc)  # for mean_cl_boot

plotDV <- function(df, dv_col, x_jitter = 0.10, y_jitter = 5) {
  
  
  ggplot(df, aes(x = rating_type, y = .data[[dv_col]], fill = machine)) +
    facet_grid(machine ~ contingency) +
    scale_y_continuous(limits = c(-10, 110), breaks = seq(0, 100, by = 10)) +
    
    # Reference line at y = 50
    geom_hline(yintercept = 50, color = "grey20", linewidth = 0.4, linetype = "dashed") +
    
    # Violin background (lighter, no outline)
    geom_violin(
      trim = TRUE, alpha = 0.25,
      color = NA
    ) +
    
    # Subject lines (jittered both horizontally & vertically)
    geom_line(
      aes(group = subj_code),
      alpha = 0.1,
      color = "grey50",
      position = position_jitter(width = x_jitter, height = y_jitter)
    ) +
    
    # Jittered points
    geom_point(
      alpha = 0.15,
      size = 1.2,
      color = "black",
      position = position_jitter(width = x_jitter, height = y_jitter)
    ) +
    
    # Bootstrapped mean ±95% CI ribbons (higher alpha)
    stat_summary(
      aes(group = machine, fill = machine),
      fun.data = mean_cl_boot,
      geom = "ribbon",
      alpha = 0.55,
      color = NA
    ) +
    
    # Mean line (per machine)
    stat_summary(
      aes(color = machine, group = machine),
      fun = mean,
      geom = "line",
      size = 1
    ) +
    
    # Mean point with thin black border
    stat_summary(
      aes(fill = machine),
      fun.data = mean_cl_boot,
      geom = "pointrange",
      shape = 21,
      color = "black",
      size = 0.9,
      stroke = 0.6
    ) +
    
    # Median point (white triangle)
    stat_summary(
      fun = median,
      geom = "point",
      shape = 24,
      size = 2.5,
      fill = "white",
      color = "black"
    ) +
    
    # Color scales
    scale_fill_manual(values = machine_colors) +
    scale_color_manual(values = machine_colors) +
    
    labs(
      x = "Structure Rating",
      y = "Rating Value"
    ) +
    
    # APA-friendly theme (no gridlines)
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 25, hjust = 1),
      axis.text.y = element_text(size = 10),
      panel.spacing = unit(3, "lines"),
      panel.grid = element_blank(),
      strip.background = element_rect(fill = "white", color = "grey70", linewidth = 0.4),
      strip.text = element_text(face = "bold", size = 12),
      axis.title = element_text(face = "bold")
    )
}

# ---- Run and Save ----
p1 <- plotDV(data_long, "rating_value")
p1

#ggsave("structure_rating_by_machine_APA_colored.svg", plot = p1, width = 8, height = 6)
#ggsave("structure_rating_by_machine_APA_colored.pdf", plot = p1, width = 8, height = 6)

#ggsave("structure_rating_by_machine_APA_colored_pilot.svg", plot = p1, width = 8, height = 6)
#ggsave("structure_rating_by_machine_APA_colored_pilot.pdf", plot = p1, width = 8, height = 6)

```

Test the difference between prior and posterior structure ratings depending on machine and contingency condition using a mixed ANOVA with contingency and machine as between-subject factors and rating_type (prior vs. posterior) as within-subject factor


```{r}
anova_data <- data %>%
  select(subj_code, machine, contingency, structure_rating_prior, structure_rating_posterior) %>%
  pivot_longer(cols = c(structure_rating_prior, structure_rating_posterior),
               names_to = "rating_type",
               values_to = "rating_value")

anova_data$rating_type <- factor(anova_data$rating_type, levels = c("structure_rating_prior", "structure_rating_posterior"),
                                labels = c("Prior", "Posterior"))
```


```{r}
library(afex)
library(emmeans)

a1 <- aov_car(rating_value ~ contingency*machine*rating_type  + Error(subj_code/rating_type), anova_data, anova_table = list(es = "pes")) # the anova model
a1 
```

repeat analysis as mixed effects model 

```{r}
library(lme4)
library(lmerTest)  # gives p-values

# drop the non-realized contingency levels 
anova_data <- droplevels(anova_data)
with(anova_data, table(machine, contingency))

model <- lmer(
  rating_value ~ machine * contingency * rating_type +
    (1 | subj_code),
  data = anova_data
)

anova(model, type = 3)


```


```{r}
emmip(
  model,
  contingency ~ rating_type | machine,
  CIs = TRUE
)
```




```{r}
# means
group_means <- emmeans(a1, c("rating_type", "contingency", "machine")) 
group_means
```
compute a contrast with emmeans to test if the prior-posterior change is larger in the opaque machine condition compared to the transparent machine condition, across all contingency conditions

```{r}
pairs(emmeans(a1, ~ rating_type|machine), adjust = "none") -> deltas

deltas


# now test the difference in these deltas between the two machine conditions
pairs(pairs(emmeans(a1, ~ rating_type*machine), adjust = "none"), adjust = "none")

```

```{r}
pairs(emmeans(a1, ~ rating_type|contingency*machine), adjust = "none") -> deltas

deltas
```


Now make a plot that shows the mean difference (together with the 95% CI) between prior and posterior structure ratings for each machine and contingency condition.

```{r}
data_diff <- data %>%
  mutate(rating_diff = structure_rating_posterior - structure_rating_prior) %>%
  group_by(machine, contingency) %>%
  summarise(
    mean_diff = mean(rating_diff, na.rm = TRUE),
    ci_lower = mean_diff - qt(0.975, df = n() - 1) * sd(rating_diff, na.rm = TRUE) / sqrt(n()),
    ci_upper = mean_diff + qt(0.975, df = n() - 1) * sd(rating_diff, na.rm = TRUE) / sqrt(n())
  )
```


Now make a plot that visualizes this 

```{r}
ggplot(data_diff, aes(x = contingency, y = mean_diff, fill = machine)) +
  #facet_wrap(~ contingency) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.7) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), 
                position = position_dodge(width = 0.9), width = 0.2) +
  scale_fill_manual(values = machine_colors) +
  labs(
    title = "Change in Structure Ratings (Posterior - Prior)",
    x = "Contingency",
    y = "Mean Difference"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 25, hjust = 1)
  ) -> p2
p2
# save plot 
#ggsave("delta.svg", plot = p2, width = 6, height = 4)
```

Compute the different between prior and posterior for each participant and append it to data frame 

```{r}
data <- data %>%
  mutate(rating_diff = structure_rating_posterior - structure_rating_prior)
```

Code for each participate what kind of rating difference occurred: no difference, increase, decrease. For the "no difference" type, allow for some noise of plus minus 5 points 

```{r}
data <- data %>%
  mutate(rating_diff_type = case_when(
    rating_diff > 0 ~ "increase",
    rating_diff < 0 ~ "decrease",
    TRUE ~ "same"
  ))
```


Now make a plot that shows the proportions of these rating_diff_type for each machine and contingency condition; as error bars, show 95% Proportion CIs (using binom.confint from the binom package)

```{r}
library(binom)
data_diff_type <- data %>%
  group_by(machine, contingency, rating_diff_type) %>%
  summarise(
    count = n()
  ) %>%
  group_by(machine, contingency) %>%
  mutate(
    total = sum(count),
    proportion = count / total
  ) %>%
  ungroup()
# compute 95% CIs for proportions
data_diff_type <- data_diff_type %>%
  rowwise() %>%
  mutate(
    ci = list(binom.confint(count, total, method = "wilson"))
  ) %>%
  unnest_wider(ci) %>%
  select(-c(method, mean, x))


ggplot(data_diff_type, aes(x = contingency, y = proportion, fill = rating_diff_type)) +
  facet_wrap(~ machine) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.7) +
  geom_errorbar(aes(ymin = lower, ymax = upper), 
                position = position_dodge(width = 0.9), width = 0.2) +
  scale_fill_manual(values = c("increase" = "#1b9e77", "same" = "#d95f02", "decrease" = "#7570b3")) +
  labs(
    title = "Proportion of Rating Change Types",
    x = "Contingency",
    y = "Proportion", 
    fill = "change type"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 25, hjust = 1)
  )+ # define y axis limits so that they go from 0 to 1 in steps of 0.25
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25))-> p3
p3
# save plot 
#ggsave("subgroups.svg", plot = p3, width = 14, height = 6)
#ggsave("subgroups.pdf", plot = p3, width = 14, height = 6)

#ggsave("subgroups_pilot.svg", plot = p3, width = 14, height = 6)
#ggsave("subgroups_pilot.pdf", plot = p3, width = 14, height = 6)
```


Determine the proportion of participants in the "transparent" condition who fell into the "same" cluster and a "rating_diff" of 0
```{r}
data %>%
  filter(machine == "transparent" & rating_diff_type == "same") %>%
  summarise(
    count = n()
  ) -> transparent_same_count

data %>%
  filter(machine == "transparent" & rating_diff_type == "same" & structure_rating_prior == 0) %>%
  summarise(
    count = n()
  ) -> transparent_same_count_zero_prior


transparent_same_count_zero_prior/transparent_same_count
```

